Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease

### 2023-11-22 doc/manuscript/translocation.qmd
Try analysis with omega on juvenile y1 survival probability

---
bibliography: translocation.bib
csl: pnas.csl
# format: docx
format :
  # latex:
  pdf:
    cite-method: natbib
    include-in-header:
      text: |
        \usepackage{blkarray}
editor: visual
editor_options:
  markdown:
    wrap: sentence
---

@@

Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability $\omega$ of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
These six populations all had observed yearly adult survival greater than 0.5.
When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).

Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of 0 to 0.32 (@fig-viability B).
This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment.
When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50-year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).

@@

$$
\begin{bmatrix}
L_1 \\
L_2 \\
L_3 \\
J_1 \\
J_2 \\
A_R \\ 
A_T
\end{bmatrix}(t + 1) = 
\begin{bmatrix}
  0 & 0 & 0 & 0 & 0 & \sigma_{A_R} p_F F & \sigma_{A_T} p_F F \\
  \sigma_{L1} p_{L1} & 0 & 0 & 0 & 0 & 0 & 0 \\
  0 & \sigma_{L2} p_{L2} & 0 & 0 & 0 & 0 & 0\\
  \sigma_{L1} (1 - p_{L1}) & \sigma_{L2} (1 - p_{L2}) & \sigma_{L3} & 0 & 0 & 0 & 0 \\
  0 & 0 & 0 & \sigma_{J1} p_{J1} & 0 & 0 & 0 \\
  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) \omega & \sigma_{J2} \omega & \sigma_{A_R} & 0 \\
  0 & 0 & 0 & 0 & 0 & 0 & \sigma_{A_T} 
\end{bmatrix} \begin{bmatrix}
L_1 \\
L_2 \\
L_3 \\
J_1 \\
J_2 \\
A_R \\
A_T \\
\end{bmatrix}(t) 
$$

The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$, and the probability of a juvenile successfully recruiting to an adult $\omega$.

@@

#### Swab Methods

@@

#### Mark-recapture model structure

We estimated survival and recruitment for each site using open population mark recapture models based on @joseph2018.
For each individual $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected. 
Capture histories of $M$ individuals are modeled, though only $N_s$ individuals were captured.
This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012parameter]. 
Here, $M$ was chosen to be three times the number of observed individuals ($3N_s$) to be considerably greater than our prior guess of $N_s$.

We denote the true state of individual $i$ as $u_{i, t}$ for primary period $t = 1,..., n_t$. 
The four states that we consider are: $u_{i, t} = 1$ for individuals that have not recruited, $u_{i, t} =2$ for live adults, and $u_{i, t} = 3$ for dead adults. 
Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$. 
Each primary period $t$ occurs within one year, but within a year there are multiple primary periods.
We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$. 
Years increment by one until the final year of the mark recapture efforts, which we denote $n_y$: $y \in \{1, 2, ..., n_y\}$.
We assume that within a primary period, the state of each individual does not change (i.e., individuals do not recruit into the adult population, gain or lose Bd infection, or die). 
This assumption is justified by the short time intervals between surveys within primary periods.

Live individuals are detected with probability $p_j$, which is modeled as:

$$p_j = \text{logit}^{-1}(X_j^{(p)} \beta^{(p)}),$$

where $X_j^{(p)}$ is a known row vector and $\beta^{(p)}$ an unknown parameter vector. 
Not recruited and dead individuals are never captured. 
We bundle these assumptions about the observation probabilities for survey $j$ into an emission matrix $\Omega_j$:

$$
\Omega_j =
\begin{blockarray}{ccc}
  \text{Not detected} & \text{Detected} \\
\begin{block}{(cc)c}
  1 & 0 & \text{Not recruited} \\
  1 - p_j & p_j & \text{Alive} \\
  1 & 0 & \text{Dead} \\
\end{block}
\end{blockarray}
$$

TODO:

- State model description
- State transition matrix for natural recruits
- State transition matrix for introduced individuals
- Parameter model description
- Priors
- Specification of the joint distribution (possible diagram of the DAG)

@@

| **Parameter**                                                             | **Value**            | **Source**                                                         |
|:-----------------------------|:---------------|:--------------------------|
| $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{L_2}$, Yearly survival probability of year 2 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{L_3}$, Yearly survival probability of year 3 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{J_1}$, Yearly survival probability of year 1 juveniles           | 0.25                 | Estimated from field data, observations, natural history knowledge |
| $\sigma_{J_2}$, Yearly survival probability of year 2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{A_R}$, Yearly survival probability of naturally recruited adults | Varies by population | Estimated from mark-recapture studies                              |
| $\sigma_{A_T}$, Yearly survival probability of translocated adults        | Varies by population | Estimated from mark-recapture studies                              |
| $p_{L_1}$, Probability of a year 1 tadpoles remaining as a tadpoles       | 1                    | Estimated from field data, observations, natural history knowledge |
| $p_{L_2}$, Probability of a year 2 tadpoles remaining as a tadpoles       | 0.25                 | Estimated from field data, observations, natural history knowledge |
| $p_{J_1}$, Probability of a year 1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations, natural history knowledge |
| $p_F$, Probability of a adult female reproducing in a year                | 0.5                  | Could be as high at 1, based on field observations                 |
| $F$, Number of surviving eggs produced by an adult female                 | 100                  | From observations of captive frogs                                 |
| $\omega$, Probability of juvenile successfully recruiting to an adult     | Varies yearly        | Explored different values of this parameter                        |

>>>>

---
bibliography: translocation.bib
csl: pnas.csl
# format: docx
format :
  latex:
    cite-method: natbib
editor: visual
editor_options:
  markdown:
    wrap: sentence
---

@@

Given observed yearly adult survival probabilities (from analysis of CMR data) and a yearly survival probability of the year 1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.XX-1.XX for these six populations).
These six populations all had observed yearly adult survival greater than 0.5.
As year 1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of eight of twelve population was greater than 1 (@fig-viability A).

Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in juvenile survival and recruitment (for year 1 and year 2 juveniles, the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average percent reduction in juvenile survival was less than 50% (given baseline year 1 juvenile survival probability of 0.25; @fig-viability B).
This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in survival and recruitment.
When we decreased baseline year 1 juvenile survival to 0.09, 50-year extinction probability was of less than 50% for four of these populations (X-X) and greater than 50% (X-X) for two of these populations (@fig-viability B).
In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \>50% regardless of baseline year 1 survival probability.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).

@@

$$
\begin{bmatrix}
L_1 \\
L_2 \\
L_3 \\
J_1 \\
J_2 \\
A_R \\ 
A_T
\end{bmatrix}(t + 1) = 
\begin{bmatrix}
  0 & 0 & 0 & 0 & 0 & \sigma_{A_R} p_F F & \sigma_{A_T} p_F F \\
  \sigma_{L1} p_{L1} & 0 & 0 & 0 & 0 & 0 & 0 \\
  0 & \sigma_{L2} p_{L2} & 0 & 0 & 0 & 0 & 0\\
  \sigma_{L1} (1 - p_{L1}) & \sigma_{L2} (1 - p_{L2}) & \sigma_{L3} & 0 & 0 & 0 & 0 \\
  0 & 0 & 0 & \sigma_{J1} p_{J1}(1 - \omega) & 0 & 0 & 0 \\
  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) (1 - \omega) & \sigma_{J2} (1 - \omega) & \sigma_{A_R} & 0 \\
  0 & 0 & 0 & 0 & 0 & 0 & \sigma_{A_T} 
\end{bmatrix} \begin{bmatrix}
L_1 \\
L_2 \\
L_3 \\
J_1 \\
J_2 \\
A_R \\
A_T \\
\end{bmatrix}(t) 
$$

The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile \$p\_{J1}\$.
The parameter $(1 - \omega)$ gives the percent reduction in baseline juvenile survival and recruitment due environmental stochasticity.

@@

@@

@@

| **Parameter**                                                             | **Value**            | **Source**                                                         |
|:----------------------------|:-----------------|:-------------------------|
| $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{L_2}$, Yearly survival probability of year 2 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{L_3}$, Yearly survival probability of year 3 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{J_1}$, Yearly survival probability of year 1 juveniles           | 0.25                 | Estimated from field data, observations, natural history knowledge |
| $\sigma_{J_2}$, Yearly survival probability of year 2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{A_R}$, Yearly survival probability of naturally recruited adults | Varies by population | Estimated from mark-recapture studies                              |
| $\sigma_{A_T}$, Yearly survival probability of translocated adults        | Varies by population | Estimated from mark-recapture studies                              |
| $p_{L_1}$, Probability of a year 1 tadpoles remaining as a tadpoles       | 1                    | Estimated from field data, observations, natural history knowledge |
| $p_{L_2}$, Probability of a year 2 tadpoles remaining as a tadpoles       | 0.25                 | Estimated from field data, observations, natural history knowledge |
| $p_{J_1}$, Probability of a year 1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations, natural history knowledge |
| $p_F$, Probability of a adult female reproducing in a year                | 0.5                  | Could be as high at 1, based on field observations                 |
| $F$, Number of surviving eggs produced by an adult female                 | 100                  | From observations of captive frogs                                 |
| $\omega$, Probability of juvenile successfully recruiting to an adult     | Varies yearly        | Explored different values of this parameter                        |
